### 2018-02-03 submission/manuscript.Rmd
update to change CRC in specific places

**Background.** An increasing body of literature suggests that there is a crucial role for the microbiota in colorectal cancer (CRC) pathogenesis. Important drivers within this context have ranged from individual microbes to the whole community. Our study expands on a recent meta-analysis investigating microbial biomarkers for CRC by testing the hypothesis that the bacterial community has important associations to both early (adenoma) and late (carcinoma) stage disease. To test this hypothesis we examined both feces (n = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and colon tissue (`r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` total samples from `r length(unique(flemer_data$Individual)) + sum(filter(tissue_n, study == "lu")[, c("control", "adenoma")])/2 + sum(filter(tissue_n, study == "burns")[, c("control", "cancer")])/2 + sum(filter(tissue_n, study == "chen")[, c("control", "cancer")])/2 + round(sum(filter(tissue_n, study == "dejea")[, c("control", "cancer")])/2) + 1 + sum(filter(tissue_n, study == "geng")[, c("control", "cancer")])/2 +  sum(filter(tissue_n, study == "sana")[, c("control", "cancer")])` individuals) across 14 previously published 16S rRNA gene sequencing studies on CRC and the microbiota.

@@

Colorectal cancer (CRC) is a growing world-wide health problem in which the microbiota has been purported to play an active role in disease pathogenesis [@siegel_cancer_2016; @flynn_metabolic_2016]. Numerous studies have shown the importance of both individual microbes [@goodwin_polyamine_2011; @abed_fap2_2016; @arthur_intestinal_2012; @kostic_fusobacterium_2013; @wu_human_2009] and the overall community [@zackular_manipulation_2016; @zackular_gut_2013; @baxter_structure_2014] in tumorgenesis using mouse models of CRC. There have also been numerous case/control studies investigating the microbiota in the formation of both adenoma and carcinoma. A recent meta-analysis investigated whether specific biomarkers could be consistently identified using multiple data sets [@shah_leveraging_2017]. This meta-analysis focused on identifying microbial signatures of CRC (biomarkers) but did so on a small total number of individuals and only investigated stool. This present meta-analysis addresses some of these major shortcomings.

Although there has been an intense focus on microbiota-based biomarker discovery for CRC, the number of candidate genera seem to be endless. Some studies point towards mouth-associated genera such as *Fusobacterium*, *Peptostreptococcus*, *Parvimonas*, and *Porphyromonas* as key enriched genera [@ahn_human_2013; @baxter_microbiota-based_2016; @chen_human_2012; @wang_structural_2012; @kostic_fusobacterium_2013; @burns_virulence_2015; @zeller_potential_2014; @flemer_tumour-associated_2017]. Yet, even in these studies, mouth-associated genera are far from the only microbes identified to be associated with CRC. These other genera include, but are not limited to, *Providencia*, *Mogibacterium*, *Enterococcus*, *Escherichia/Shigella*, *Klebsiella*, and *Streptococcus* [@burns_virulence_2015; @chen_human_2012; @wang_structural_2012]. In fact, there is good *in vivo* evidence that *Escherichia/Shigella* and *Streptococcus* can be important in the pathogenesis of CRC [@ecoli_Arthur_2014; @arthur_intestinal_2012; @strep_Aymeric_2017]. Other studies have also identified *Akkermansia muciniphila* and *Bacteroides fragilis* as potential markers of CRC with good mechanistic studies for the latter [@weir_stool_2013; @wang_structural_2012; @bfrag_Boleij_2014]. A recent meta-analysis confirmed the correlations of certain mouth-associated genera and *Akkermansia muciniphila* with carcinoma [@shah_leveraging_2017]. However, the sample size (n = 509) is equal to or less than some more recent individual studies investigating the microbiota and CRC, making it hard to know how extrapolatable these findings are. That particular meta-analysis also added more potential microbial associations to both carcinoma (*Pantoea agglomerans* *Ruminococcus*, *Lactobacillus*) and adenoma (*Prevotella*, *Methanosphaera*, *Succinovibrio*, *Haemophilus parainfluenzae*, *Ruminococcus*, *Lactobacillus*) stages of CRC that need to be investigated further, since a number of these genera have been found to be enriched in controls and not disease [@baxter_microbiota-based_2016; @zeller_potential_2014; @burns_virulence_2015]. Additionally, genera like *Roseburia* have been found in some studies to be increased in CRC but in others to either be decreased or have no difference [@geng_diversified_2013; @wang_structural_2012; @flemer_tumour-associated_2017; @dejea_microbiota_2014].

Most of these studies have focused on carcinoma but the adenoma observations are not any clearer at identifying candidate genera of disease. Groups focusing on broad scale community metrics have found that metrics such as richness are decreased in the adenoma stage of CRC versus controls. Other studies have identified *Lactococcus*, *Pseudomonas*, *Acidovorax*, *Cloacibacterium*, *Helicobacter*, *Lactobacillus*, *Bilophila*, *Desulfovibrio*, and *Mogibacterium* to be increased in adenoma [@sanapareddy_increased_2012; @lu_mucosal_2016; @hale_shifts_2017]. Based on the studies mentioned, there seems to be very little overlap between the genera identified to be associated with adenoma and carcinoma, with *Lactobacillus* being one of the few commonalities.

Targeting the identification of CRC microbial biomarkers within stool seems logical since it offers an easy and cost-effective way to stratify risk and the current gold standard for diagnosis, a colonoscopy, can be time-consuming and is not without risk of complications. Although stool represents an easy and less invasive way to assess risk, it is not clear how well this sample reflects adenoma- and carcinoma- associated microbial communities. Some studies have tried to assess this in health and disease but are limited by their sample size [@flemer_tumour-associated_2017; @Flynn_preprint_2017]. Sampling the microbiota directly associated with colon tissue may provide clearer answers but is not without limitations. The community present for sampling following the colonoscopy bowel prep may reflect the better adhered microbiota versus the resident community. Additionally, these samples contain more host DNA, potentially limiting the types of analysis that can be done. It is well known that low biomass samples can be very difficult to work with and results can be study dependent due to the randomness of contamination [@Salter_contamination_2014].

In comparison to the previous meta-analysis, this study significantly increases the total stool samples investigated, re-examines important genera across adenoma and carcinoma across study, and examines differences and similarities between stool and tissue microbiota in the context of CRC. Importantly, this analysis and approach could provide valuable insights into the common genera that are both protective and detrimental in CRC and whether broad bacterial community measurements can account for these changes that were not provided by earlier meta-analysis studies [@shah_leveraging_2017].

Using both feces (n = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and colon tissues (`r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` samples from `r length(unique(flemer_data$Individual)) + sum(filter(tissue_n, study == "lu")[, c("control", "adenoma")])/2 + sum(filter(tissue_n, study == "burns")[, c("control", "cancer")])/2 + sum(filter(tissue_n, study == "chen")[, c("control", "cancer")])/2 + round(sum(filter(tissue_n, study == "dejea")[, c("control", "cancer")])/2) + 1 + sum(filter(tissue_n, study == "geng")[, c("control", "cancer")])/2 +  sum(filter(tissue_n, study == "sana")[, c("control", "cancer")])` individuals) totaling over `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T) + sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` total samples across 14 studies [@hale_shifts_2017; @flemer_tumour-associated_2017; @chen_human_2012; @wang_structural_2012; @lu_mucosal_2016; @brim_microbiome_2013; @weir_stool_2013; @dejea_microbiota_2014; @sanapareddy_increased_2012; @ahn_human_2013; @burns_virulence_2015; @zeller_potential_2014; @baxter_microbiota-based_2016; @geng_diversified_2013] [Table 1 & 2], we expand both the breadth and scope of the previous meta-analysis to investigate whether the bacterial community or specific members are more important risk factors for both adenoma and carcinoma stages of CRC. To accomplish this we first assessed whether bacterial diversity changes throughout disease (control to adenoma to carcinoma) and if it results in an increased relative risk (RR) for adenoma or carcinoma stages of CRC. We then assessed what genera, if any, increase or decrease the RR of adenoma or carcinoma stages of CRC. Next, using Random Forest models, we analyzed whether the full community or only the combined top 5 increased and top 5 decreased RR genera resulted in better model classification, based on the area under the curve (AUC). Finally, we also examined at what effect and sample size the studies used were powered for and the sample size needed to get to the traditionally accepted 80% power. Our results from these analyses suggests that the bacterial community changes as disease severity worsens, that specific members are important for disease classification, and that many of the studies are underpowered for assessing small effect sizes.

@@

***Lower Bacterial Diversity is Associated with Increased RR of CRC:***
To assess differences in broad scale community metrics as disease severity worsens Operational Taxonomic Unit (OTU) richness, evenness, and Shannon diversity measurements were power transformed and Z-score normalized. These metrics are commonly used to assess the total number of OTUs, the equality of their abundance, and the overall diversity, respectively. Using linear mixed-effect models to control for study and variable region we assessed whether OTU richness, evenness, or Shannon diversity changed in a step-wise manner with disease severity. In stool, there was a significant decrease in both evenness and Shannon diversity as disease severity moved from control to adenoma to carcinoma (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannoneven")[, "BH"], digits = 2, nsmall = 2))` and `r paste(format(filter(stool_lme, alpha_metric == "shannon")[, "BH"], digits = 2, nsmall = 2))`, respectively) [Figure 1A]. We next tested whether the detectable differences in community significantly increased in RR of having an adenoma or carcinoma. For fecal samples, a decrease versus the overall median in evenness resulted in a significantly increased RR for carcinoma  (RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`),  P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) and adenoma (RR = `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) while a decrease versus the overall median in Shannon diversity only increased the RR for carcinoma (RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2, nsmall = 2))`) [Figure 2]. Using the Bray-Curtis distance metric and PERMANOVA, it was also possible to identify significant bacterial community changes, in specific studies, for both carcinoma-associated and adenoma-associated microbiota versus control [Table S1 & S2].

@@

***Mouth-Associated Genera are Associated with an Increased RR of CRC:***
Next, we asked if being higher than the median relative abundance, for any specific genera, resulted in an altered RR for adenoma or carcinoma, in stool and colon tissue, due to our previous observations of small increases in RR using OTU richness and Shannon diversity. To investigate this we analyzed all common genera across each study, in colon tissue or stool, and assessed whether a relative abundance higher than the median results in an increase or decrease in RR. Mouth-associated genera were commonly found in the top 5 genera associated with an increased RR of having an adenoma (*Pyramidobacter* [Figure 3A] and *Rothia* [Figure 3C]) and carcinoma (*Fusobacterium*, *Parvimonas*, *Porphyromonas*, and *Peptostreptococcus* [Figure 3B] and *Fusobacterium* [Figure 3D]) for both stool and colon tissue samples. Conversely, genera commonly associated with a normal gastrointestinal tract were correlated with a decreased RR for both adenoma and carcinoma for both stool and colon tissue samples [Figure 3]. Even though mouth-associated genera were identified across disease stage, there was little direct overlap of the top 5 increased or decreased RR genera between both stages and sample site.

When observing RRs with a P-value less than 0.05 there was almost no overlap between genera from stool or colon tissue and when they were similar the RR was in opposite directions (e.g. *Lactococcus*) [Table S8 & S9]. Many of the genera that had RRs with a P-value under 0.05 for colon tissue are also highly prevalent in contamination, specifically, *Novosphingobium*, *Selemonas*, and *Achromobacter* [Figure 3 & Table S8-S9]. For carcinoma stage of CRC, certain mouth-associated genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*) had a high RR for both colon tissue and stool samples [Table S10 & S11]. Finally, these data suggest that the most significantly increased RR genera for tissue was *Camplyobacter* while in stool it was *Peptostreptococcus* [Table S10 & S11].

***Select Genera Models can Recapitulate Whole Community Models:***
Since specific genera increased RR for carcinoma over diversity metrics we assessed whether the bacterial community was better at classifying disease versus only a select group of genera. We selected these genera based on their RR and P-value significance and used two approaches to test this question.
The first approach used genus level data and tested for differences in AUC between all genera and selected genera. A single study was used for training the model prior to testing on all other studies and this was repeated for every study in the meta-analysis. The second approach used OTU level data and tested for a generalized decrease in the 10-fold cross validation (CV) model AUC which is a common approach used to guard against over-fitting. This was applied across study and the AUC of the all OTUs model was compared against the model that used only OTUs that taxonomically classified to selected genera.

@@

Our study identifies clear differences in diversity, both at the community level and for individual genera, present in patients with and without CRC [Figure 1-3]. Although there was a step-wise decrease in diversity as disease progressed from control to adenoma to carcinoma, this did not translate into large effect sizes for the RR of CRC. Even though mouth-associated genera increased the RR of having a carcinoma, they did not consistently increase the RR of having an adenoma. Additionally, our observations suggest that by combining mouth-associated and CRC protective microbes we can classify either adenoma or carcinoma stage of disease as well as models that use the full community.

@@

The colon tissue-based studies did not provide a clearer understanding of how the microbiota may be associated with CRC. Generally, the full OTU-based models of unmatched and matched colon tissue samples were concordant with stool samples showing that GI resident microbes were the most prevalent in the top 10 most important variables across study [Figure S7E & F]. Unlike in stool, *Fusobacterium* was the only mouth-associated bacteria consistently present in the top 10 most important variables of the full carcinoma stage models [Figure S7B-C & E-F]. The majority of the colon tissue-based results seem to be study specific with many of the top 10 taxa being present only in a single study. Additionally, the presence of genera associated with contamination, within the top 10 most important variables for the genera and OTU models is worrying. The low bacterial biomass of tissue samples coupled with potential contamination could explain why these results seem to be more sporadic than the stool results.

@@

By aggregating together a large collection of studies analyzing both fecal and colon tissue samples, we are able to provide evidence supporting the importance of the bacterial community in CRC. Further, the data presented here suggests that mouth-associated microbes can gain a foothold within the colon and are are commonly associated with the greatest RR of having a carcinoma. No conclusive signal with these mouth-associated microbes could be detected for adenoma stage of disease. Our observations also highlight the importance of power and sample number considerations when investigating the microbiota and adenoma stage of disease due to the subtle changes in the community. Overall, associations of microbiota with the carcinoma stage of CRC are much stronger than those with the adenoma stage.

@@

***Matched versus Unmatched Tissue Samples:***
In general, tissue samples with control and CRC samples from different individuals were classified as unmatched while samples that belonged to the same individual were classified as matched. Studies with matched data included Burns, Dejea, Geng, and Lu while those with unmatched data were from Burns, Flemer, Chen, and Sanapareddy. For some studies samples became unmatched when a corresponding matched sample did not make it through sequence processing. All samples, from both tissue sample types, were analyzed together for the linear mixed-effect models with samples from the same individual corrected for. For all other analysis, not mentioned herein, matched and unmatched samples were analyzed separately using the statistical approaches mentioned in the Statistical Analysis section.

>>>>

**Background.** An increasing body of literature suggests that there is a crucial role for the microbiota in colorectal cancer (CRC) pathogenesis. Important drivers within this context have ranged from individual microbes to the whole community. Our study expands on a recent meta-analysis investigating microbial biomarkers for tumors by testing the hypothesis that the bacterial community has important associations to both early (adenoma) and late (carcinoma) stage disease. To test this hypothesis we examined both feces (n = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and colon tissue (`r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` total samples from `r length(unique(flemer_data$Individual)) + sum(filter(tissue_n, study == "lu")[, c("control", "adenoma")])/2 + sum(filter(tissue_n, study == "burns")[, c("control", "cancer")])/2 + sum(filter(tissue_n, study == "chen")[, c("control", "cancer")])/2 + round(sum(filter(tissue_n, study == "dejea")[, c("control", "cancer")])/2) + 1 + sum(filter(tissue_n, study == "geng")[, c("control", "cancer")])/2 +  sum(filter(tissue_n, study == "sana")[, c("control", "cancer")])` individuals) across 14 previously published 16S rRNA gene sequencing studies on CRC and the microbiota.

@@

Colorectal cancer (CRC) is a growing world-wide health problem in which the microbiota has been purported to play an active role in disease pathogenesis [@siegel_cancer_2016; @flynn_metabolic_2016]. Numerous studies have shown the importance of both individual microbes [@goodwin_polyamine_2011; @abed_fap2_2016; @arthur_intestinal_2012; @kostic_fusobacterium_2013; @wu_human_2009] and the overall community [@zackular_manipulation_2016; @zackular_gut_2013; @baxter_structure_2014] in tumorgenesis using mouse models of CRC. There have also been numerous case/control studies investigating the microbiota in the formation of both adenoma and carcinoma. A recent meta-analysis investigated whether specific biomarkers could be consistently identified using multiple data sets [@shah_leveraging_2017]. This meta-analysis focused on identifying microbial signatures of tumors (biomarkers) but did so on a small total number of individuals and only investigated stool. This present meta-analysis addresses some of these major shortcomings.

Although there has been an intense focus on microbiota-based biomarker discovery for tumors, the number of candidate genera seem to be endless. Some studies point towards mouth-associated genera such as *Fusobacterium*, *Peptostreptococcus*, *Parvimonas*, and *Porphyromonas* as key enriched genera [@ahn_human_2013; @baxter_microbiota-based_2016; @chen_human_2012; @wang_structural_2012; @kostic_fusobacterium_2013; @burns_virulence_2015; @zeller_potential_2014; @flemer_tumour-associated_2017]. Yet, even in these studies, mouth-associated genera are far from the only microbes identified to be associated with CRC. These other genera include, but are not limited to, *Providencia*, *Mogibacterium*, *Enterococcus*, *Escherichia/Shigella*, *Klebsiella*, and *Streptococcus* [@burns_virulence_2015; @chen_human_2012; @wang_structural_2012]. In fact, there is good *in vivo* evidence that *Escherichia/Shigella* and *Streptococcus* can be important in the pathogenesis of CRC [@ecoli_Arthur_2014; @arthur_intestinal_2012; @strep_Aymeric_2017]. Other studies have also identified *Akkermansia muciniphila* and *Bacteroides fragilis* as potential markers of CRC with good mechanistic studies for the latter [@weir_stool_2013; @wang_structural_2012; @bfrag_Boleij_2014]. A recent meta-analysis confirmed the correlations of certain mouth-associated genera and *Akkermansia muciniphila* with carcinoma [@shah_leveraging_2017]. However, the sample size (n = 509) is equal to or less than some more recent individual studies investigating the microbiota and colorectal tumors, making it hard to know how extrapolatable these findings are. That particular meta-analysis also added more potential microbial associations to both carcinoma (*Pantoea agglomerans* *Ruminococcus*, *Lactobacillus*) and adenoma (*Prevotella*, *Methanosphaera*, *Succinovibrio*, *Haemophilus parainfluenzae*, *Ruminococcus*, *Lactobacillus*) stages of disease that need to be investigated further, since a number of these genera have been found to be enriched in controls and not disease [@baxter_microbiota-based_2016; @zeller_potential_2014; @burns_virulence_2015]. Additionally, genera like *Roseburia* have been found in some studies to be increased in tumors but in others to either be decreased or have no difference [@geng_diversified_2013; @wang_structural_2012; @flemer_tumour-associated_2017; @dejea_microbiota_2014].

Most of these studies have focused on carcinoma but the adenoma observations are not any clearer at identifying candidate genera of disease. Groups focusing on broad scale community metrics have found that metrics such as richness are decreased in the adenoma stage of disease versus controls. Other studies have identified *Lactococcus*, *Pseudomonas*, *Acidovorax*, *Cloacibacterium*, *Helicobacter*, *Lactobacillus*, *Bilophila*, *Desulfovibrio*, and *Mogibacterium* to be increased in adenoma [@sanapareddy_increased_2012; @lu_mucosal_2016; @hale_shifts_2017]. Based on the studies mentioned, there seems to be very little overlap between the genera identified to be associated with adenoma and carcinoma, with *Lactobacillus* being one of the few commonalities.

Targeting the identification of tumor microbial biomarkers within stool seems logical since it offers an easy and cost-effective way to stratify risk and the current gold standard for diagnosis, a colonoscopy, can be time-consuming and is not without risk of complications. Although stool represents an easy and less invasive way to assess risk, it is not clear how well this sample reflects adenoma- and carcinoma- associated microbial communities. Some studies have tried to assess this in health and disease but are limited by their sample size [@flemer_tumour-associated_2017; @Flynn_preprint_2017]. Sampling the microbiota directly associated with colon tissue may provide clearer answers but is not without limitations. The community present for sampling following the colonoscopy bowel prep may reflect the better adhered microbiota versus the resident community. Additionally, these samples contain more host DNA, potentially limiting the types of analysis that can be done. It is well known that low biomass samples can be very difficult to work with and results can be study dependent due to the randomness of contamination [@Salter_contamination_2014].

In comparison to the previous meta-analysis, this study significantly increases the total stool samples investigated, re-examines important genera across adenoma and carcinoma across study, and examines differences and similarities between stool and tissue microbiota in the context of colorectal tumors. Importantly, this analysis and approach could provide valuable insights into the common genera that are both protective and detrimental in CRC and whether broad bacterial community measurements can account for these changes that were not provided by earlier meta-analysis studies [@shah_leveraging_2017].

Using both feces (n = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and colon tissues (`r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` samples from `r length(unique(flemer_data$Individual)) + sum(filter(tissue_n, study == "lu")[, c("control", "adenoma")])/2 + sum(filter(tissue_n, study == "burns")[, c("control", "cancer")])/2 + sum(filter(tissue_n, study == "chen")[, c("control", "cancer")])/2 + round(sum(filter(tissue_n, study == "dejea")[, c("control", "cancer")])/2) + 1 + sum(filter(tissue_n, study == "geng")[, c("control", "cancer")])/2 +  sum(filter(tissue_n, study == "sana")[, c("control", "cancer")])` individuals) totaling over `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T) + sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` total samples across 14 studies [@hale_shifts_2017; @flemer_tumour-associated_2017; @chen_human_2012; @wang_structural_2012; @lu_mucosal_2016; @brim_microbiome_2013; @weir_stool_2013; @dejea_microbiota_2014; @sanapareddy_increased_2012; @ahn_human_2013; @burns_virulence_2015; @zeller_potential_2014; @baxter_microbiota-based_2016; @geng_diversified_2013] [Table 1 & 2], we expand both the breadth and scope of the previous meta-analysis to investigate whether the bacterial community or specific members are more important risk factors for both adenoma and carcinoma stages of disease. To accomplish this we first assessed whether bacterial diversity changes throughout disease (control to adenoma to carcinoma) and if it results in an increased relative risk (RR) for adenoma or carcinoma stages of disease We then assessed what genera, if any, increase or decrease the RR of adenoma or carcinoma stages of disease Next, using Random Forest models, we analyzed whether the full community or only the combined top 5 increased and top 5 decreased RR genera resulted in better model classification, based on the area under the curve (AUC). Finally, we also examined at what effect and sample size the studies used were powered for and the sample size needed to get to the traditionally accepted 80% power. Our results from these analyses suggests that the bacterial community changes as disease severity worsens, that specific members are important for disease classification, and that many of the studies are underpowered for assessing small effect sizes.

@@

***Lower Bacterial Diversity is Associated with Increased RR of Tumors:***
To assess differences in broad scale community metrics as disease severity worsens Operational Taxonomic Unit (OTU) richness, evenness, and Shannon diversity measurements were power transformed and Z-score normalized. These metrics are commonly used to assess the total number of OTUs, the equality of their abundance, and the overall diversity, respectively. Using linear mixed-effect models to control for study and variable region we assessed whether OTU richness, evenness, or Shannon diversity changed in a step-wise manner with disease severity. In stool, there was a significant decrease in both evenness and Shannon diversity as disease severity moved from control to adenoma to carcinoma (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannoneven")[, "BH"], digits = 2, nsmall = 2))` and `r paste(format(filter(stool_lme, alpha_metric == "shannon")[, "BH"], digits = 2, nsmall = 2))`, respectively) [Figure 1A]. We next tested whether the detectable differences in community significantly increased in RR of having an adenoma or carcinoma. For fecal samples, a decrease versus the overall median in evenness resulted in a significantly increased RR for carcinoma  (RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`),  P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) and adenoma (RR = `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) while a decrease versus the overall median in Shannon diversity only increased the RR for carcinoma (RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2, nsmall = 2))`) [Figure 2]. Using the Bray-Curtis distance metric and PERMANOVA, it was also possible to identify significant bacterial community changes, in specific studies, for both carcinoma-associated and adenoma-associated microbiota versus control [Table S1 & S2].

@@

***Mouth-Associated Genera are Associated with an Increased RR of Tumor:***
Next, we asked if being higher than the median relative abundance, for any specific genera, resulted in an altered RR for adenoma or carcinoma, in stool and colon tissue, due to our previous observations of small increases in RR using OTU richness and Shannon diversity. To investigate this we analyzed all common genera across each study, in colon tissue or stool, and assessed whether a relative abundance higher than the median results in an increase or decrease in RR. Mouth-associated genera were commonly found in the top 5 genera associated with an increased RR of having an adenoma (*Pyramidobacter* [Figure 3A] and *Rothia* [Figure 3C]) and carcinoma (*Fusobacterium*, *Parvimonas*, *Porphyromonas*, and *Peptostreptococcus* [Figure 3B] and *Fusobacterium* [Figure 3D]) for both stool and colon tissue samples. Conversely, genera commonly associated with a normal gastrointestinal tract were correlated with a decreased RR for both adenoma and carcinoma for both stool and colon tissue samples [Figure 3]. Even though mouth-associated genera were identified across disease stage, there was little direct overlap of the top 5 increased or decreased RR genera between both stages and sample site.

When observing RRs with a P-value less than 0.05 there was almost no overlap between genera from stool or colon tissue and when they were similar the RR was in opposite directions (e.g. *Lactococcus*) [Table S8 & S9]. Many of the genera that had RRs with a P-value under 0.05 for colon tissue are also highly prevalent in contamination, specifically, *Novosphingobium*, *Selemonas*, and *Achromobacter* [Figure 3 & Table S8-S9]. For carcinoma stage of disease, certain mouth-associated genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*) had a high RR for both colon tissue and stool samples [Table S10 & S11]. Finally, these data suggest that the most significantly increased RR genera for tissue was *Camplyobacter* while in stool it was *Peptostreptococcus* [Table S10 & S11].

***Select Community Models can Recapitulate Whole Community Models:***
Since specific genera increased RR for carcinoma over diversity metrics we assessed whether the bacterial community was better at classifying disease versus only a select group of genera. We selected these genera based on their RR and P-value significance and used two approaches to test this question.
The first approach used genus level data and tested for differences in AUC between all genera and selected genera. A single study was used for training the model prior to testing on all other studies and this was repeated for every study in the meta-analysis. The second approach used OTU level data and tested for a generalized decrease in the 10-fold cross validation (CV) model AUC which is a common approach used to guard against over-fitting. This was applied across study and the AUC of the all OTUs model was compared against the model that used only OTUs that taxonomically classified to selected genera.

@@

Our study identifies clear differences in diversity, both at the community level and for individual genera, present in patients with and without CRC [Figure 1-3]. Although there was a step-wise decrease in diversity as disease progressed from control to adenoma to carcinoma, this did not translate into large effect sizes for the RR of tumors. Even though mouth-associated genera increased the RR of having a carcinoma, they did not consistently increase the RR of having an adenoma. Additionally, our observations suggest that by combining mouth-associated and CRC protective microbes we can classify either adenoma or carcinoma stage of disease as well as models that use the full community.

@@

The colon tissue-based studies did not provide a clearer understanding of how the microbiota may be associated with tumors. Generally, the full OTU-based models of unmatched and matched colon tissue samples were concordant with stool samples showing that GI resident microbes were the most prevalent in the top 10 most important variables across study [Figure S7E & F]. Unlike in stool, *Fusobacterium* was the only mouth-associated bacteria consistently present in the top 10 most important variables of the full carcinoma stage models [Figure S7B-C & E-F]. The majority of the colon tissue-based results seem to be study specific with many of the top 10 taxa being present only in a single study. Additionally, the presence of genera associated with contamination, within the top 10 most important variables for the genera and OTU models is worrying. The low bacterial biomass of tissue samples coupled with potential contamination could explain why these results seem to be more sporadic than the stool results.

@@

By aggregating together a large collection of studies analyzing both fecal and colon tissue samples, we are able to provide evidence supporting the importance of the bacterial community in CRC. Further, the data presented here suggests that mouth-associated microbes can gain a foothold within the colon and are are commonly associated with the greatest RR of having a carcinoma. No conclusive signal with these mouth-associated microbes could be detected for adenoma stage of disease. Overall, associations between the microbiota and carcinoma stage of disease were much stronger than with the adenoma stage. Our observations also highlight the importance of power and sample number considerations when investigating the microbiota and adenoma stage of disease due to the subtle changes in the community.

@@

***Matched versus Unmatched Tissue Samples:***
In general, tissue samples with control and tumor samples from different individuals were classified as unmatched while samples that belonged to the same individual were classified as matched. Studies with matched data included Burns, Dejea, Geng, and Lu while those with unmatched data were from Burns, Flemer, Chen, and Sanapareddy. For some studies samples became unmatched when a corresponding matched sample did not make it through sequence processing. All samples, from both tissue sample types, were analyzed together for the linear mixed-effect models with samples from the same individual corrected for. For all other analysis, not mentioned herein, matched and unmatched samples were analyzed separately using the statistical approaches mentioned in the Statistical Analysis section.
